Profitability Prediction of Consulting Projects in the Construction Industry

### 2016-08-25 finalwriteup/draft.rmd
done!

Predicting how long it will take a person to do something is a notoriously tricky task. People make mental calculations of this sort every day - estimating how long it will take to drive to meet someone on time, how many tasks you can check off in a day, or how long it will take to cook a dish. As most people have experienced, these estimates can be particularly inaccurate if the task is new and difficult. These errors are trivial in everyday activities, but for consulting businesses that solve complex problems, similar time estimation errors determine their financial well-being. Furthermore,

Consulting businesses give expert advice to other professionals in exchange for a fee based on the amount of time their consulting staff spends on the project. The deliverable 'expert advice' is generally in the form of a document or a package of documents. A document could range from a 5-page report written by a single person to a sequence of hundreds of drawings assembled by a team of 10 over many years. Despite the range of possible deliverables, before any project commences, the client and consultant must agree on a fee, or at least a fee structure. The fee structure can be as creative as the engaged parties wish, but in fact tend to fit a few common models. A couple of examples are:

@@

This thesis focuses on the risk taken by a consulting company when offering a fixed fee. Clients commonly seek fixed price quotes from several consultants before selecting one or negotiating further. Quoted fees must be competitive.

The way a consulting manager calculates a fixed price varies from industry to industry and even company to company. Typically a consulting manager has experience in the type of project they are quoting and, after reviewing the project details, can use a combination of intuition, comparison to past projects, and rules of thumb (such as a set fraction of the entire client budget for a large project). Such qualitative methods can result in profit margins, which ultimately depend on the skill (or luck) the manager has in predicting the amount of time a project demands.

This thesis focuses on a single case study consulting company in the construction industry that has persevered through many fixed price projects that resulted in losses. A source of insight into the loss making projects was identified - twelve years of passively collected data that described each project in terms of their clients, invoice history, employee hours and technical details. Was it possible to find trends in what makes a project loss making and can these characteristics be identified before a fixed price is offered? Can these trends be modeled into a predictive algorithm, and if so, how could the outcomes be integrated into business decisions? Finally, if the predictive outcomes were integrated into decision-making, would they definitely improve the business' bottom line? These questions are explored in depth and answered over the course of this dissertation.

@@

Groundbreaking methods into calculating construction costs by @Ma2014 mobilised data from building information modeling (BIM) models, although this is still technically a detailed analysis of the costs. The idea is to quantify the cost of building a structure directly from a three dimensional BIM model created by the design consultants (engineers and architects). Ma and Liu's [-@Ma2014] first trial was to automate the cost calculation of a reinforced concrete structure. They programmed their algorithm to intelligently establish construction techniques for each element, as architectural or engineering models do not provide this. From the construction technique and material, a cost was derived for each element. This was successful and @Ma2014 aim to further their work so that more construction information can be intelligently obtained from BIM models. A system like Ma and Liu's [-@Ma2014] would definitely accelerate the cost estimation process and reduce human error, however it is still an 'inside view' and could still deliver inaccurate estimates similar to detailed analysis by a person. The detailed view does not 'step back' and take into account set backs in construction that may have affected past similar projects. There is potential for this innovation to be combined with an 'outside view' system.

@@

Similar to the construction industry, expert judgment or detailed analysis, is the most widely practised method for effort estimation [@Shepperd1996; @Molokken2003]. This is despite many years of research being dedicated to developing algorithmic models that, in the research context, outperform expert judgment. A survey by @Moores1992 found that software tools assisting a software project were in significant use, but only the project planning tools, not the cost estimating tools. This suggests a mental link for project managers to price jobs based on *details* as opposed to an outside view. Another study by @Heemstra1992 found there was no evidence estimation accuracy improved when estimation tools were used. @Bergeron1992 similarly found jobs that used algorithmic models were actually associated with less accurate estimates. However, this trend may be coincidental due to a lack of cases where estimation tools were used [@Molokken2003]. The scarcity of real project evidence that algorithmic estimation tools improve estimations and an unintuitive mental jump could be reasons contributing to detailed expert analysis remaining the most widespread technique and mirrors what was found in the construction industry.

@@

Clearly, Neural Networks can mimick a flexible range of relationships, and if more nodes are added, more flexibility is available [@Putler2012]. A previously mentioned disadvantage is that the algorithm is a black box because the internal structure is too complex for interpretation. They also require a lot of training data relative to other methods. Despite these disadvantages, Neural Networks have found success in high-level tasks where the model structure does not need to be understood. These include hand writing recognition, vehicle control, and face recognition [@Haykin2004].

@@

In ANOVA, categorical and continuous explanatory variables are treated differently. For categorical variables, differences in the mean values and variances of the response variable in each group are tested. In contrast, linear relationships are examined between continuous explanatory variables and the response variable. The continuous variables were normalised to kerb the effects of possible high leverage points, and logarithmic or cube root transformations were used for this. Trials of various variable interactions revealed that interactions between:

@@

The two regression models predicting 'return per dollar' resulted in high mean RMSE's ranging from \$`r mean(results.rf.core$final.RMSE) %>% round(2)` to \$`r mean(results.A2$final.RMSE) %>% round(2)`. The next section discusses possible reasons for the poor results and the decision to halt development of regression models.

@@

@@

Random Forests and Boosted Trees passively interact explanatory variables, but for Logistic Regression, these must be manually input. Only interaction terms with statistically significant p-values from the Logisitic Regression were included in order to reduce noise.

@@

Only three of the model blending techniques used the 8 variables determined above. These were FWLS, Random Forest, and Boosted Trees. The remaining 3 techniques used only the predictions of the three individual models. All six methods were compared against the original Logistic Regression model. Initially 100 train/test runs were performed for each blending method by running 20 iterations of 5-fold cross validation. The AUC results of these iterations are displayed below:

@@

The simple Logistic Regression blend and simple averaging of the individual models (Logistic Regression, Boosted Trees and Random Forests) gave virtually the same distribution of results, where both AUC distributions were significantly greater than the original Logistic Regression model. The p-values from the two-sample t-tests were `r pvals$p_value[1]` and `r pvals$p_value[2]` for simple average and simple Logistic Regression methods respectively. It was encouraging that blended models produced higher AUC values than the original Logistic Regression model, however for the case study business, the real measure of success is the model's impact on the business' overall profits. Higher AUC values did not guaruntee more lucrative profit curves so several methods, including the original individual models as well as blended methods, were progressed to the profit analysis.

@@

The plot below illustrates the distrubtion of profit curves for each method, where the solid lines join the mean values at each threshold point. The grey ribbon illustrates the 95% confidence interval for the profit improvement ratio at each threshold point for the 100 models.

@@

```{r, fig.height = 5.5, fig.cap = c("Distribution of projects according to the probability outputs from a typical simple Logistic Regression blended model. A bargraph of counts and relative profits are shown.")}

@@

The general aim was to use statistical techniques to model the profitability of projects for a case study consulting business using their internal CRM data. This was rigorously completed by approaching the prediction of 'profitability' in two ways: as a regression problem predicting the degree of profitability and as a binary classification problem predicting either profit or loss. Several statistical and machine learning approaches were applied to both problems including Naive Bayes, Bayesian Networks, Linear Regression, Random Forests, gradient Boosted Trees as well as multiple methods of blending the individual models' output.

@@

Lastly, hypothesis 2 proposed that the best predictive model developed from hypothesis 1 could be shown to improve the overall profitability (bottom line) of the case study business. This theory was tested using a range of predictive models because AUC scores were not directly linked to profitability once the models were applied to a business decision scenario. The models output probability scores, so an optimal probability threshold needed to be found that would dictate whether a job was classified as 'profitable' or 'loss making'. A range of threshold values were trialed, and a decision framework was created where if a project was scored below the threshold it was accepted while projects above the threshold were rejected. If a project was accepted, the profits and losses were summed while if a project was rejected the profits and losses were forfeited. A perfect model would reject all loss making jobs and accept all profitable jobs, thereby improving overall profits. Final results showed the simple Logistic Regression blend of the individual Logistic Regression, Random Forest and Boosted Tree models improved profits the most. The 95% confidence interval for these improvements was between `r best.fin %>% filter(method == "simp.log") %>% .$lower_bound_95CI - 100`% and `r best.fin %>% filter(method == "simp.log") %>% .$upper_bound_95CI - 100`% using a probability threshold of `r best.t` (approximately `r ((simp.l %>% filter(mean.iter>best.t) %>% nrow)/nrow(simp.l)*100) %>% round(1)`% of projects).

These results contribute significantly to the research in cost estimation in three ways: the applied methods, the decision framework, and the appeal to user trust. Ensemble tree methods and blending had been applied minimally to cost estimation previously, and past research had favoured Linear Regression, Neural Networks and CBR. Ensemble tree methods are advantageous because they provide insight into the model's structure while predicting at a similar level to Neural Networks. Next, a clear framework for how a business could tangibly improve profits was developed and proven successful. Previous studies have verified predictive accuracy but stopped short of how the algorithm would affect decisions and what the measured benefits would be. Lastly, the decision framework was progressed even further into a prototype application where user trust was addressed. Past surveys have reported that despite decades of promising research in cost estimation, decision maker's still mainly use an 'inside view' during fee estimation. This was addressed by displaying intuitive plots of the model structure as well as a Nearest Neighbour graph of similar internal projects that provide an 'outside view'. Further work is required to test user confidence in the output.

Another topic identified for future research was to test how well managers estimate some numeric project input variables. In particular, time span, team size, total invoiced amount, and percentage of hours completed by professionals should be tested as these variables were calculated post project completion. Time span and total invoiced amount were discretised into wide categories, which should be easier for a manager to choose between. Initial estimates of input variables could also be kept in the database for future analysis.

Overall, this work has successfully built a mathematical blend of Logistic Regression, Random Forests, and Boosted Tree models, built on a consulting company's internal project data. This blended model can predict whether a project will be profitable or not and in a reasonable decision framework, can guide managers in rejecting financially risky projects and improving profitability of the business.

>>>>

Predicting how long it will take a person to do something is a notoriously tricky task. People make mental calculations of this sort every day - estimating how long it will take to drive to meet someone on time, how many tasks you can check off in a day, or how long it will take to cook a dish. As most people have experienced, these estimates can be particularly inaccurate if the task is new and difficult. These errors are trivial in everyday activities, but for consulting businesses that solve complex problems, similar time estimation errors determine their financial well-being.

Consulting businesses give expert advice to other professionals in exchange for a fee based on the amount of time their consulting staff spends on the project. The deliverable 'expert advice' is generally in the form of a document or a package of documents. A document could range from a 5-page report written by a single person to a sequence of hundreds of drawings assembled by a team of 10 over many years. Despite the range of possible deliverables, before any project commences, the client and consultant must agree on a fee, or at least a fee structure. The fee structure can be as creative as the engaged parties wish, but in fact tend to fit a few common models. Two examples are:

@@

This thesis focuses on the risk taken by a consulting company when offering a fixed fee. Clients commonly seek fixed price quotes from several consultants before selecting one or negotiating further. Hence, quoted fees must be competitive.

The way a consulting manager calculates a fixed price varies from industry to industry and even company to company. Typically a consulting manager has experience in the type of project they are quoting and, after reviewing the project details, can use a combination of intuition, comparison to past projects, and rules of thumb (such as a set fraction of the entire client budget for a large project). Such qualitative methods can result in profit margins, which ultimately depend on the skill (or luck) of the manager in predicting the amount of time a project demands.

This thesis focuses on a single case study consulting company in the construction industry that has persevered through many fixed price projects that resulted in losses. A source of insight into the loss making projects was identified - twelve years of passively collected data that described each project in terms of their clients, invoice history, employee hours and technical details. Was it possible to find trends in what makes a project loss making and can these characteristics be identified before a fixed price is offered? Can these trends be modeled into a predictive algorithm, and if so, how could the outcomes become integrated into business decisions? Finally, if the predictive outcomes were integrated into decision-making, would they definitely improve the business' bottom line? These questions are explored in depth and answered over the course of this dissertation.

@@

Groundbreaking methods into calculating construction costs by @Ma2014 mobilised data from building information modeling (BIM) models, although this is still technically a detailed analysis of the costs. The idea is to quantify the cost of building a structure directly from a three dimensional BIM model created by the design consultants (engineers and architects). Ma and Liu's [-@Ma2014] first trial was to automate the cost calculation of a reinforced concrete structure. They programmed their algorithm to intelligently establish construction techniques for each element, as architectural or engineering models do not provide this. From the construction technique and material, a cost was derived for each element. This was successful and @Ma2014 aim to further their work so that more construction information can be intelligently obtained from BIM models. A system like Ma and Liu's [-@Ma2014] would definitely accelerate the cost estimation process and reduce human error, however it is still an 'inside view' and could still deliver inaccurate estimates similar to detailed analysis by a person. The detailed view does not 'step back' and take into account setbacks in construction that may have affected past similar projects. There is potential for this innovation to be combined with an 'outside view' system.

@@

Similar to the construction industry, expert judgment or detailed analysis, is the most widely practiced method for effort estimation [@Shepperd1996; @Molokken2003]. This is despite many years of research being dedicated to developing algorithmic models that, in the research context, outperform expert judgment. A survey by @Moores1992 found that software tools assisting a software project were in significant use, but only the project planning tools, not the cost estimating tools. This suggests a mental link for project managers to price jobs based on *details* as opposed to an outside view. Another study by @Heemstra1992 found there was no evidence estimation accuracy improved when estimation tools were used. @Bergeron1992 similarly found jobs that used algorithmic models were actually associated with less accurate estimates. However, this trend may be coincidental due to a lack of cases where estimation tools were used [@Molokken2003]. The scarcity of real project evidence that algorithmic estimation tools improve estimations and an unintuitive mental jump could be reasons contributing to detailed expert analysis remaining the most widespread technique and mirrors what was found in the construction industry.

@@

Clearly, Neural Networks can mimic a flexible range of relationships, and if more nodes are added, more flexibility is available [@Putler2012]. A previously mentioned disadvantage is that the algorithm is a black box because the internal structure is too complex for interpretation. They also require a lot of training data relative to other methods. Despite these disadvantages, Neural Networks have found success in high-level tasks where the model structure does not need to be understood. These include hand writing recognition, vehicle control, and face recognition [@Haykin2004].

@@

In ANOVA, categorical and continuous explanatory variables are treated differently. For categorical variables, differences in the mean values and variances of the response variable in each group are tested. In contrast, linear relationships are examined between continuous explanatory variables and the response variable. The continuous variables were normalised to curb the effects of possible high leverage points, and logarithmic or cube root transformations were used for this. Trials of various variable interactions revealed that interactions between:

@@

The two regression models predicting 'return per dollar' resulted in high mean RMSE's ranging from \$`r mean(results.rf.core$final.RMSE) %>% round(2)` to \$`r mean(results.A2$final.RMSE) %>% round(2)`. Example individual results from the ANOVA Linear Regression and Random Forest regression models can be found in the Appendix, section 11.1. The next section discusses possible reasons for the poor results and the decision to halt development of regression models.

@@

Summaries of an example model from each of the individual classification methods can be found in the Appendix, section 11.2.

@@

Random Forests and Boosted Trees passively interact explanatory variables, but for Logistic Regression, these must be manually input. Only interaction terms with statistically significant p-values from the Logistic Regression were included in order to reduce noise.

@@

Only three of the model blending techniques used the 8 variables determined above. These were FWLS, Random Forest, and Boosted Trees. The remaining 3 techniques used only the predictions of the three individual models. All six methods were compared against the original Logistic Regression model. Initially 100 train/test runs were performed for each blending method by running 20 iterations of 5-fold cross validation. Summaries of a single example model from each of the blending methods can be found in the Appendix, section 11.3. The AUC results of these iterations are displayed below:

@@

The simple Logistic Regression blend and simple averaging of the individual models (Logistic Regression, Boosted Trees and Random Forests) gave virtually the same distribution of results, where both AUC distributions were significantly greater than the original Logistic Regression model. The p-values from the two-sample t-tests were `r pvals$p_value[1]` and `r pvals$p_value[2]` for simple average and simple Logistic Regression methods respectively. It was encouraging that blended models produced higher AUC values than the original Logistic Regression model, however for the case study business, the real measure of success is the model's impact on the business' overall profits. Higher AUC values did not guarantee more lucrative profit curves so several methods, including the original individual models as well as blended methods, were progressed to the profit analysis.

@@

The plot below illustrates the distribution of profit curves for each method, where the solid lines join the mean values at each threshold point. The grey ribbon illustrates the 95% confidence interval for the profit improvement ratio at each threshold point for the 100 models.

@@

```{r, fig.height = 5.5, fig.cap = c("Distribution of projects according to the probability outputs from a typical simple Logistic Regression blended model. A bar graph of counts and relative profits are shown.")}

@@

The general aim was to use statistical techniques to predict the profitability of projects for a case study consulting business using their internal CRM data. This was rigorously completed by approaching the prediction of 'profitability' in two ways: as a regression problem predicting the degree of profitability and as a binary classification problem predicting either profit or loss. Several statistical and machine learning approaches were applied to both problems including Naive Bayes, Bayesian Networks, Linear Regression, Random Forests, gradient Boosted Trees as well as multiple methods of blending the individual models' output.

@@

Lastly, hypothesis 2 proposed that predictive models developed from hypothesis 1 could be shown to improve the overall profitability (bottom line) of the case study business. A range of probability threshold values were trialed for each method using a decision framework where projects scored below the threshold were accepted, while projects above the threshold were rejected. If a project was rejected, the profits and losses were forfeited, and the remaining accepted profits and losses were summed. Final results showed the simple Logistic Regression blend of the individual Logistic Regression, Random Forest and Boosted Tree models improved profits the most. The 95% confidence interval for these improvements was between `r best.fin %>% filter(method == "simp.log") %>% .$lower_bound_95CI - 100`% and `r best.fin %>% filter(method == "simp.log") %>% .$upper_bound_95CI - 100`% using a probability threshold of `r best.t` (approximately `r ((simp.l %>% filter(mean.iter>best.t) %>% nrow)/nrow(simp.l)*100) %>% round(1)`% of projects).

These results contribute significantly to the research in cost estimation in three ways: the applied methods, the decision framework, and the appeal to user trust. Ensemble tree methods and blending had been applied minimally to cost estimation previously, even though ensemble trees provide insight into model structure while predicting at a similar level to Neural Networks. Next, previous studies have verified predictive accuracy but stopped short of how the algorithm would affect decisions and what the measured benefits would be. This study presented a clear framework for how a business could improve profits by applying the algorithm. Lastly, a prototype application was developed where insight into the model structure encouraged user trust  and a Nearest Neighbour chart presented an ‘outside view’ based on the company’s projects. Past surveys have reported that despite decades of promising research in cost estimation, decision maker's still mainly use an 'inside view' during fee estimation. Further work is required to test user confidence in the output.

Another topic identified for future research was to test how well managers estimate some numeric project input variables. In particular, time span, team size, total invoiced amount, and percentage of hours completed by professionals should be tested as these variables were calculated post project completion. Time span and total invoiced amount were discretised into wide categories, which should be easier for a manager to choose between.

Overall, this work has successfully built a mathematical blend of Logistic Regression, Random Forests, and Boosted Tree models, from a consulting company's internal project data. This blended model can predict whether a project will be profitable or not and in a reasonable decision framework, can guide managers in rejecting financially risky projects and improving profitability of the business.

# Appendix

Example results from a single model from each method. This is broken into the three main types of methods that were trialled:

1. Regression models
2. Classification models
3. Blended classification models

## Regression models

ANOVA Linear Regression and Random Forests were applied to profitability as a regression problem.

### ANOVA regression

Example summary of model built with core variables only:
```{r}

all8a = read.csv("all8a.csv")[,-1]

#function below outputs summary of model
final.output<- function(df=all8a, no.tests = 50, core = TRUE) {

        summary.tests = matrix(NA, nrow = no.tests, ncol = 3)

        for(i in 1:no.tests){

        if(core == FALSE){
        samp = sample.int(nrow(df), 1)                        
        #create sample df - use all variables that are complete from a sample row of data
        sample = df[samp,]
        sample = sample[,!is.na(sample)]
        vars= names(sample)
        vars = vars[! vars %in% c('mlsto','Year','return.pdol',
                                  'code.client', 'code.contact')]
        variables = vars
        } else{
                variables = c('inv.mlsto.log', 'Discipline', 'timespan.cbrt',
                              'no.users', 'pc.pro', 'Business')
        }

        #make reduced dataset with complete cases for all variables we want to use
        reduced = df[complete.cases(df[,variables]),]

        formula = paste("return.pdol ~ ")

        for(j in 1:length(variables)){
                formula = paste(formula, variables[j], sep= " + ")
        }

        if('inv.mlsto.log' %in% variables &
           'pc.pro' %in% variables){
                formula = paste(formula, 'inv.mlsto.log*pc.pro', sep= " + ")
        }

        if('inv.mlsto.log' %in% variables &
           'code.client' %in% variables){
                formula = paste(formula, 'inv.mlsto.log*code.client', sep= " + ")
        }

        if('JD.Second' %in% variables &
           'no.users' %in% variables){
                formula = paste(formula, 'JD.Second*no.users', sep= " + ")
        }

        samp_size = floor(0.75*nrow(reduced))
        create.train = base::sample.int(nrow(reduced), samp_size)
        train.df = reduced[create.train,]
        test.df = reduced[-create.train,]

        cats = c('code.client', 'code.contact', 'JD.Second', 'Business', 'majority.pos')
        for(k in seq_along(cats)){
                if(cats[k] %in% variables){
                test.df = test.df[test.df[,cats[k]] %in% unique(train.df[,cats[k]]),]
        }
        }

        final = aov(as.formula(formula), data = train.df)

        fit= summary(final)

        return(fit)
}

final.output(df=all8a, no.tests = 1, core = TRUE)

Example summary of model built with all variables. Note that each model used a subset of all variables based on the available variables for a randomly selected project:

final.output(df=all8a, no.tests = 1, core = FALSE)

### Random Forest

Example summary of model built with core variables only:

final.rf.output<- function(df=all8a, no.tests = 50, core = TRUE) {

        summary.tests = matrix(NA, nrow = no.tests, ncol = 3)

        for(i in 1:no.tests){

        if(core == FALSE){
        samp = sample.int(nrow(df), 1)                        
        #create sample df - use all variables that are complete from a sample row of data
        sample = df[samp,]
        sample = sample[,!is.na(sample)]
        vars= names(sample)
        vars = vars[! vars %in% c('mlsto','Year','return.pdol')]
        variables = vars
        } else{
                variables = c('inv.mlsto.log', 'Discipline', 'timespan.cbrt',
                              'no.users', 'pc.pro', 'Business')
        }

        #make reduced dataset with complete cases for all variables we want to use
        reduced = df[complete.cases(df[,variables]),]

        if('code.client' %in% variables){
        sample = factor(sample(unique(reduced$code.client),size=53,
                                     replace=FALSE))
        reduced= reduced[reduced$code.client %in% sample,]

        reduced$code.client= factor(reduced$code.client)
        }

        if('code.contact' %in% variables){
        sample1 = factor(sample(unique(reduced$code.contact),size=53,
                                     replace=FALSE))
        reduced= reduced[reduced$code.contact %in% sample1,]

        reduced$code.contact= factor(reduced$code.contact)
        }

        formula = paste("return.pdol ~ ")

        for(j in 1:length(variables)){
                formula = paste(formula, variables[j], sep= " + ")
        }

        samp_size = floor(0.75*nrow(reduced))
        create.train = base::sample.int(nrow(reduced), samp_size)
        train.df = reduced[create.train,]
        test.df = reduced[-create.train,]

        cats = c('code.client', 'code.contact', 'JD.Second', 'Business', 'majority.pos')
        for(k in seq_along(cats)){
                if(cats[k] %in% variables){
                test.df = test.df[test.df[,cats[k]] %in% unique(train.df[,cats[k]]),]
        }
        }

        final = randomForest(as.formula(formula), data= train.df, mtry=5,ntree=500, importance=FALSE)

        }
        return(final)
}

rf.core <- final.rf.output(df=all8a, no.tests = 1, core = TRUE)

print(rf.core)
getTree(rf.core, 1) %>% head(20)
```

All variables:

```{r}
rf.all <- final.rf.output(df=all8a, no.tests = 1, core = FALSE)

print(rf.all)
getTree(rf.all, 1) %>% head(20)

## Classification models

Logistic Regression, Random Forests, Boosted Trees, Naïve Bayes, and Bayesian Networks were applied to profitability prediction as a classification problem (will a project be profitable or not?)

Example summary output imputed data:

# load imputed data
all10mice_f = read.csv("all10mice_May16.csv")[,-1]

#function for running models
test.roc <- function(dfs = c('all10mice_f'), seed = 100, mtrys = c(5), method = "log",
                     extra.var = NULL, mult = 1, x.folds = 10){

        auc<- rep(list(rep(0, length(dfs))), length(dfs))
        colours <- topo.colors(x.folds*mult)

        for(i in 1:length(dfs)){
                df = get(dfs[i])

                # mlsto is spinning errors, get rid of
                df = df[, !(names(df) %in% 'mlsto')]
                set.seed(seed)

                # create f.rpdol
                df$f.rpdol <- as.factor(df$b.rpdol)
                levels(df$f.rpdol)[levels(df$f.rpdol) == "0"] <- "profit"
                levels(df$f.rpdol)[levels(df$f.rpdol) == "1"] <- "loss"

                #set up 10 fold cv using caret function
                folds <- createFolds(df$f.rpdol, k = x.folds, list = T, returnTrain = F)
                #now have folds[[1]] through folds[[10]] list of rows to exclude

                #for Appendix just want one fold
                folds <- folds[[1]]

                if(mult > 1){
                        for(k in 2:mult){
                                set.seed(sample(1:1000,1))
                                folds.temp = createFolds(df$f.rpdol, k = x.folds, list = T, returnTrain = F)
                                folds = append(folds, folds.temp)
                        }
                }

                #if extra.var not found in names(df), change to NULL
                if(!is.null(extra.var)){
                        if(!(extra.var %in% names(df))) extra.var = NULL
                }

                #define formula for algorith methods, either rf, log, boost

                if(nrow(df) == 2364){
                        if(is.null(extra.var)){
                                formula <-"f.rpdol ~ Discipline + pc.pro + b.timespan.cbrt + no.users + b.inv.log + client.totinv.log + Business + majority.pos + pc.majpos.log + JD.Second + Billing.Type"

                        }}

                if(method %in% 'boost') {
                        #no JD.Second
                        formula <-"b.rpdol ~ Discipline + pc.pro + b.timespan.cbrt + no.users + b.inv.log + client.totinv.log + Business + majority.pos + pc.majpos.log"

                if(method %in% 'naiive') {
                        formula<- "b.rpdol ~ Discipline + pc.pro + b.timespan.cbrt + no.users + inv.mlsto.log + client.totinv.log + Business + majority.pos + JD.Second + pc.majpos.log + Billing.Type + JD.Second"

                        #turn predictors and response into 'ordered' variables for roc() to work
                        test<- df[folds,]
                        levels(test$f.rpdol) <- c('profit', 'loss')
                        test$f.rpdol <- ordered(test$f.rpdol)

                        if(method %in% 'rf'){
                                fit <- randomForest(as.formula(formula), data = df[-folds,], 
                                            mtry = mtrys[i], ntree = 1000)
                                pred <- predict(fit, test, type = "prob")
                                pred <- pred[,2]

                        if(method %in% 'log'){
                                fit<- glm(as.formula(formula), 
                                          data = df[-folds,], family = binomial())
                                pred <- predict(fit, test,
                                                type = "response")
                        } 
                        if(method %in% 'boost'){
                                fit <- gbm(as.formula(formula), data = df[-folds,],
                                           distribution = "bernoulli", n.trees = 10000,
                                           shrinkage = 0.001, interaction.depth = 5,
                                           n.minobsinnode = 20)

                                pred <- predict(fit, test, n.tree = 10000, type = "response")
                        }

                        if(method %in% 'naiive'){
                                fit<- naiveBayes( as.formula(formula), df[-folds,])
                                pred <- predict(fit, test, type = 'raw')
                                pred<- pred[,2]
                        }

                        #pROC
                        pred.p <- pROC::roc(test$f.rpdol, pred)

                        if (i == 1) {
                        plot.roc(pred.p, col = colours[i], print.thres = F)

                        } else{
                        plot.roc(
                        pred.p, col = colours, add = TRUE , print.thres = F)
                        }

                        auc<- auc(pred.p)
                        names(auc) = method

        assign("auc.func", auc, envir = .GlobalEnv)
        assign("formula", formula, envir = .GlobalEnv)
        return(fit)

Log.Reg <- test.roc(dfs = c('all10mice_f'), seed = 100, mtrys = c(5), method = "log",
                     extra.var = NULL, mult = 1, x.folds = 10)
auc.func

summary(Log.Reg)

### Random Forest

Rand.F <- test.roc(dfs = c('all10mice_f'), seed = 100, mtrys = c(5), method = "rf",
                     extra.var = NULL, mult = 1, x.folds = 10)
auc.func

Rand.F
summary(Rand.F)

### Boosted Trees

Boost.T <- test.roc(dfs = c('all10mice_f'), seed = 100, mtrys = c(5), method = "boost",
                     extra.var = NULL, mult = 1, x.folds = 10)
auc.func

# Summary of the first tree in the boosted tree
pretty.gbm.tree(Boost.T) %>% kable

summary(Boost.T)

Naive <- test.roc(dfs = c('all10mice_f'), seed = 100, mtrys = c(5), method = "naiive",
                     extra.var = NULL, mult = 1, x.folds = 10)
auc.func

Naive

### Bayesian Network

[@Genie]

6 different methods for blending were applied. Three were simple and 3 were complex as listed below:
Simple:

* Averaging
* Logistic Regression
* Boosted Trees

Complex:

* Logistic Regression
* Boosted Trees 
* Random Forests

### Simple Average

if(.Platform$OS.type == 'windows'){

        source("C:/Users/n9232371/Documents/github/consult/functions.R")
} else{

        source("/Users/yam/Documents/github/consult/functions.R")

df.blend <- readRDS("dfblend.rds")

iter.auc <- function(mult =1, x.folds.iter = 5, 
                     methods = c("simp.log"),
                     df.iter = df.blend) {

        #create test and train set: 0.75/0.25
        pred.list <- vector(mode = "list", length = length(methods))
        names(pred.list) = methods

        #create model averaged column
        if("average" %in% methods){
        df.iter = mutate(df.iter, model.av = (log + rf + boost)/3) 
        }

        #set up 10 fold cv using caret function
        folds <- createFolds(df.iter$f.rpdol, k = x.folds.iter, list = T, returnTrain = F)
        #now have folds[[1]] through folds[[10]] list of rows to exclude
        folds <- folds[[1]]

                train = df.iter[-folds,]
                test = df.iter[folds,]

                if ("simp.log" %in% methods) {
                        print("simp.log")
                        fit <- glm(f.rpdol ~ norm.log + norm.rf + norm.boost,
                                   data = train,
                                   family = binomial())

                        pred <- predict(fit, test, type = "response")
                        pred.list[["simp.log"]] = rbind(pred.list[["simp.log"]],
                                                        data.frame("index" = folds,
                                                                   "pred" = pred,
                                                                   "iteration" = paste("iter",ceiling(x.folds.iter), sep = "")))

                if ("simp.boost" %in% methods) {
                        print("simp.boost")
                        fit <- gbm(b.rpdol ~ norm.log + norm.rf + norm.boost, data = train,
                                   distribution = "bernoulli", n.trees = 6000,
                                   shrinkage = 0.0005, interaction.depth = 6,
                                   n.minobsinnode = 10)

                        pred <- predict(fit, test, n.tree = 6000, type = "response")
                        pred.list[["simp.boost"]] = rbind(pred.list[["simp.boost"]],
                                                        data.frame("index" = folds,
                                                                   "pred" = pred,
                                                                   "iteration" = paste("iter",ceiling(x.folds.iter), sep = "")))

                if ("orig.log" %in% methods) { 
                        print("orig.log")
                        pred <- test$log
                        pred.list[["orig.log"]] = rbind(pred.list[["orig.log"]],
                                                        data.frame("index" = folds,
                                                                   "pred" = pred,
                                                                   "iteration" = paste("iter",ceiling(x.folds.iter), sep = "")))

                if ("orig.rf" %in% methods) { 
                        print("orig.rf")
                        pred <- test$rf
                        pred.list[["orig.rf"]] = rbind(pred.list[["orig.rf"]],
                                                        data.frame("index" = folds,
                                                                   "pred" = pred,
                                                                   "iteration" = paste("iter",ceiling(x.folds.iter), sep = "")))

                if ("orig.boost" %in% methods) { 
                        print("orig.boost")
                        pred <- test$boost
                        pred.list[["orig.boost"]] = rbind(pred.list[["orig.boost"]],
                                                        data.frame("index" = folds,
                                                                   "pred" = pred,
                                                                   "iteration" = paste("iter",ceiling(x.folds.iter), sep = "")))

                if ("average" %in% methods) {
                        print("average")
                        pred <- test$model.av
                        pred.list[["average"]] = rbind(pred.list[["average"]],
                                                        data.frame("index" = folds,
                                                                   "pred" = pred,
                                                                   "iteration" = paste("iter",ceiling(x.folds.iter), sep = "")))
                }

                if ("FWLS" %in% methods) {
                        print("FWLS")
                        fit <- glm(f.rpdol ~ norm.log + norm.rf + norm.boost + b.timespan.cbrt*norm.log + b.inv.log*norm.log + b.inv.log*norm.rf + Business*norm.boost + JD.Second + no.users,
                                   data = train,
                                   family = binomial())
                        pred <- predict(fit, test, type = "response")
                        pred.list[["FWLS"]] = rbind(pred.list[["FWLS"]],
                                                        data.frame("index" = folds,
                                                                   "pred" = pred,
                                                                   "iteration" = paste("iter",ceiling(x.folds.iter), sep = "")))
                }

                if ("comp.boost" %in% methods) {
                        print("comp.boost")
                        fit <- gbm(b.rpdol ~ b.timespan.cbrt + no.users + b.inv.log + Business + JD.Second + norm.log + norm.rf + norm.boost, 
                                   data = train,
                                   distribution = "bernoulli", n.trees = 6000,
                                   shrinkage = 0.0005, interaction.depth = 3,
                                   n.minobsinnode = 10)

                        pred <- predict(fit, test, n.tree = 6000, type = "response")
                        pred.list[["comp.boost"]] = rbind(pred.list[["comp.boost"]],
                                                        data.frame("index" = folds,
                                                                   "pred" = pred,
                                                                   "iteration" = paste("iter",ceiling(x.folds.iter), sep = "")))
                }

                if ("comp.rf" %in% methods) {
                        print("comp.rf")
                        fit <- randomForest(f.rpdol ~ b.timespan.cbrt + no.users + b.inv.log + Business + JD.Second + norm.log + norm.rf + norm.boost,
                                            data = train, mtry = 3, 
                                            ntree = 1000, importance = TRUE
                        )
                        pred <- predict(fit, test, type = "prob")
                        pred <- pred[, 2]
                        pred.list[["comp.rf"]] = rbind(pred.list[["comp.rf"]],
                                                        data.frame("index" = folds,
                                                                   "pred" = pred,
                                                                   "iteration" = paste("iter",ceiling(x.folds.iter), sep = "")))
                }

        #pROC
                        pred.p <- pROC::roc(test$f.rpdol, pred)

                        plot.roc(pred.p, col = "darkblue", print.thres = F)

                        auc<- auc(pred.p)  
                        assign("auc.func", auc, envir = .GlobalEnv)

        return(fit)

simp.av <- iter.auc(mult =1, x.folds.iter = 5, 
                     methods = c("average"),
                     df.iter = df.blend)
auc.func

#### Simple Blend

simp.LR <- iter.auc(mult =1, x.folds.iter = 5, 
                     methods = c("simp.log"),
                     df.iter = df.blend)
auc.func
summary(simp.LR)

#### Complex Logistic Regression Blend

```{r}
complex.LR <- iter.auc(mult =1, x.folds.iter = 5, 
                     methods = c("FWLS"),
                     df.iter = df.blend)
auc.func
summary(complex.LR)

### Boosted Trees

#### Simple Blend

```{r}
simp.boost <- iter.auc(mult =1, x.folds.iter = 5, 
                     methods = c("simp.boost"),
                     df.iter = df.blend)
auc.func
summary(simp.boost)

# Summary of the first tree in the boosted tree
pretty.gbm.tree(simp.boost) %>% kable

#### Complex Blend
